---
title: "Reproducible Research Project 1"
author: "Keith Krause"
date: "10/11/2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Loading and Processing the Activity Data

## Load Packages
```{r packages}
library(lubridate)
library(dplyr)
library(ggplot2)
```
## Load Activity Data from Working Directory
```{r Activity}
activity<-read.csv("activity.csv", stringsAsFactors = FALSE)
```
## Explore the activity data
```{r Explore Data}
summary(activity)
str(activity)
head(activity)
```
## Process and transform data
```{r date}
activity$date<-ymd(activity$date)
```
# 2. Histogram of Total Number of Steps Taken per Day
## Caluculate total steps per day and plot histogram
```{r steps per day}
stepsperday<-tapply(activity$steps, activity$date, sum)
hist(stepsperday, xlab="Number of Steps", main = "Steps per Day")
```

# 3. Calculate mean and median of Total Number of Steps per Day
```{r mean median}
meanperday<- mean(stepsperday, na.rm = TRUE)
medianperday<- median(stepsperday, na.rm=TRUE)
```
Mean= 10766.19
Median=10765

# 4. Time Series Plot of Average Number of Steps
```{r Average Steps}
stepsperinterval<- tapply(activity$steps, activity$interval, mean, na.rm=TRUE)
plot(as.numeric(names(stepsperinterval)), 
     stepsperinterval,
     xlab = "Interval",
     ylab = "Steps", 
     main = "Average Number of Steps per Interval",
     type = "l")
```



# 5. The 5-minute interval that, on average, contains the maximum number of steps
```{r Max interval}
maxinterval<- names(sort(stepsperinterval, decreasing = TRUE)[1])
maxsteps<- sort(stepsperinterval, decreasing = TRUE)[1]
```
The 5-minute interval with the maximum number of steps is 835 with 206.1698 steps

# 6. Code to desctibe and show a strategy for imputing missing data
```{r Missing Values}
numberNA<- sum(is.na(activity$steps))
```
Missing Values (NA) in the data set are 2304

## Strategy for filling in all of the missing values in the data set
I will fill in the missing values (NA) in the activity data set by calculating the mean number of steps across all of the days that have data for that interval.  A new data set will be created called clean which will be used to better estimate a complete set of data.  
```{r Clean Data Set}
#split by interval
activitysplit<- split(activity, activity$interval)
#fill in missing data
for(i in 1:length(activitysplit)){
  activitysplit[[i]]$steps[is.na(activitysplit[[i]]$steps)]<- stepsperinterval[i]
}
clean<- do.call("rbind", activitysplit)
clean<- clean[order(clean$date),]
```
# 7. Histogram of Total Number of Steps Taken per day after missing values are imputed
```{r Plot Imputed Total Steps}
stepsperdayclean<-tapply(clean$steps, clean$date, sum)
hist(stepsperdayclean, xlab="Number of Steps", main="Steps per Day (Imputed Data)")
```
## Mean and Median of Total Number of Steps Taken per day with Imputed data
```{r Imputed Mean and Median}
meanstepsperdayclean<- mean(stepsperdayclean,na.rm=TRUE)
medianstepsperdayclean<- median(stepsperdayclean, na.rm = TRUE)
```
The mean for the imputed data remained the same at 10766.19.
The median for the imputed data is 10766.19 which increased by 1.19 from the original data's median calculation(10765)

# 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
## Loop to create day of the week column to classify weekend or weekday
```{r Weekday or Weekend}
for(i in 1:nrow(clean)) {
  if (weekdays(clean$date [i]) =="Saturday" | weekdays(clean$date[i])=="Sunday"){
    clean$dayofweek[i]= "weekend"
  }else {clean$dayofweek[i]="weekday"
  } 
}
```
## Calculate average steps per interval for weekend and weekdays
```{r Mean Avg Steps Weekend/Weekday}
weekendsteps<- tapply(clean[clean$dayofweek == "weekend",]$steps, clean[clean$dayofweek == "weekend",]$interval, mean, na.rm=TRUE)
weekdaysteps<- tapply(clean[clean$dayofweek == "weekday",]$steps, clean[clean$dayofweek == "weekday",]$interval, mean, na.rm=TRUE)
```
## Pannel Plot comparing Average Number of Steps Across Weekdays and Weekends
```{r Plot Avg Steps Weekend/Weekday}
par(mfrow=c(2,1))
#Weekday plot
plot(as.numeric(names(weekdaysteps)),
     weekdaysteps,
     xlab = "Interval",
     ylab = "Steps",
     main = "Activity Pattern (Weekdays)",
     type = "l")
#Weekend Plot
plot(as.numeric(names(weekendsteps)),
     weekendsteps,
     xlab = "Interval",
     ylab = "Steps",
     main = "Activity Pattern (Weekends)",
     type = "l")
```
